Calendar-based graphics for visualizing people's daily schedules

### 2018-10-08 main.Rmd
reformatted to R Journal style

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualizing people's daily schedules"
  plain: "Calendar-based graphics for visualizing people's daily schedules"
abstract: >
  Calendars are broadly used in society to display temporal information, and events. This paper describes a new \proglang{R} package with functionality to organize and display temporal data, collected on sub-daily resolution, into a calendar layout. The function \code{frame\_calendar} uses linear algebra on the date variable to restructure data into a format lending itself to calendar layouts. The user can apply the grammar of graphics to create plots inside each calendar cell, and thus the displays synchronize neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior in Melbourne, Australia, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the \proglang{R} package \pkg{sugrrants}.
keywords:
  # at least one keyword must be supplied
  formatted: [data visualization, statistical graphics, time series, grammar of graphics, "\\proglang{R}"]
  plain:     [data visualization, statistical graphics, time series, grammar of graphics, R]
preamble: >
  \usepackage{amsmath}
output: 
  bookdown::pdf_book:
    base_format: rticles::jss_article
  citation_package: biblatex
bibliography: "references.bib"
---

@@

```{r ped-map, fig.width = 4, out.width = "0.55\\linewidth", fig.cap = "(ref:ped-map-cap)"}
```

(ref:ped-map-cap) Map of the Melbourne city area with purple dots indicating sensor locations.

```{r time-series-plot, fig.height = 6, fig.cap = "(ref:time-series-plot-cap)"}
```

(ref:time-series-plot-cap) Time series plots showing the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne. Colored by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distinctive across these 3 sensors. There is an eye-catching spike which occurred at the State Library, caused by the annual White Night event on 20th of February.

```{r facet-time, fig.height = 6, fig.cap = "(ref:facet-time-cap)"}
```

(ref:facet-time-cap) Hourly pedestrian counts for 2016 faceted by sensors and days of the week using lines. It features at least two types of seasons---time of day and day of week---across all the sensors. The White Night effect on increasing pedestrian counts can be seen on Saturday in February at the State Library.

@@

We propose a new algorithm to go beyond the calendar-based heatmap. The approach is developed with three conditions in mind: (1) to display time-of-day variation in addition to longer temporal components such as day-of-week and day-of-year; (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout; (3) to enable overlaying plots consisting of multiple time series. The proposed algorithm has been implemented in the \code{frame_calendar} function in the \pkg{sugrrants} package using \proglang{R} [@R-base].

The remainder of the paper is organized as follows. Section \ref{sec:algorithm} demonstrates the construction of the calendar layout in depth. Section \ref{sec:opt} lists and describes the options that come with the \code{frame_calendar} function. Section \ref{sec:examples} presents some variations of its usage. Section \ref{sec:discussion} discusses the advantages and disadvantages of the method.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the monthly calendar over the year 2016. This is achieved by the \code{frame_calendar} function, which computes the coordinates on the calendar for the input data variables. These can then be plotted using the usual \pkg{ggplot2} package [@R-ggplot2] functions. All of the grammar of graphics [@wilkinson2006grammar; @wickham2009ggplot2] can be applied.

```{r flinders-2016, fig.height = 6.5, fig.cap = "(ref:flinders-2016-cap)"}
```

(ref:flinders-2016-cap) The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of the data into a $3 \times 4$ monthly grid represents all the traffic in 2016. The disparities between week day and weekend along with public holiday are immediately apparent.

@@

```{r month-diagram, out.width = "360pt", out.height = "250pt", fig.cap = "(ref:month-diagram-cap)"}
include_graphics("figure/month.pdf")
```

(ref:month-diagram-cap) Illustration of the indexing layout for cells in a month, where $k$ is day of the week, $g$ is day of the month, $(i, j)$ indicates grid position.

@@

```{r year-diagram, out.width = "360pt", out.height = "250pt", fig.cap = "(ref:year-diagram-cap)"}
include_graphics("figure/year-diagram.pdf")
```

(ref:year-diagram-cap) Illustration of the indexing layout for months of one year, where $M$ and $N$ indicate number of rows and columns, $b$ is a space parameter separating cells.

@@

The algorithm has several optional parameters that modify the layout, direction of display, scales, plot size and switching to polar coordinates. These are accessible to the user by the inputs to the function  \code{frame_calendar}:

@@

```{r flinders-polar, fig.height = 6.5, fig.cap = "(ref:flinders-polar-cap)"}
```

(ref:flinders-polar-cap) Figure \ref{fig:flinders-2016} in circular layout, which is referred to as star plots. The daily periodicity on work days are clearly visible.

@@

By default, global scaling is done for values in each plot, with the global minimum and maximum used to fit values into each cell. If the emphasis is comparing trend rather than magnitude, it is useful to scale locally. For temporal data this would harness the temporal components. The choices include: free scale within each cell (\code{free}), cells derived from the same day of the week (\code{free_wday}), or cells from the same day of the month (\code{free_mday}). The scaling allows for the comparisons of absolute or relative values, and the emphasis of different temporal variations.

@@

The \code{free_wday} scales each week day together. It can be useful to comparing trends across week days, allowing relative patterns for weekends versus week days to be examined. Similarly, the \code{free_mday} uses free scaling for any day within a given month.

```{r flinders-free, fig.height = 6.5, fig.cap = "(ref:flinders-free-cap)"}
```

(ref:flinders-free-cap) Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive, however it is impossible to compare the size of peaks between days.

@@

```{r overlay, fig.height = 7, fig.cap = "(ref:overlay-cap)"}
```

(ref:overlay-cap) Overlaying line graphs of the 3 sensors in the monthly calendar. Flagstaff station is not as busy as the other two on non-work days.

@@

```{r facet, fig.height = 11, fig.cap = "(ref:facet-cap)"}
```

(ref:facet-cap) Line charts, embedded in the $6 \times 2$ monthly calendar, colored and faceted by the 3 sensors. The variations of an individual sensor are emphasised, and the shapes can be compared across the cells and sensors.

@@

The \code{frame_calendar} function is not constrained to line plots. The full range of plotting capabilities in \pkg{ggplot2} is essentially available. Figure \ref{fig:scatterplot} shows a lag scatterplot at Flinders Street Station, where the current hourly count is assigned to the \code{x} argument and the lagged hourly count to the \code{y} argument. This figure is organized in the daily calendar layout. Figure \ref{fig:scatterplot} indicates two primary patterns, strong autocorrelation on weekends, and weaker autocorrelation on work days. At the higher counts, on week days, the next hour sees possibly substantial increase or decrease in counts, essentially revealing a bimodal distribution of consecutive counts, as supported by Figure \ref{fig:flinders-2016}.

```{r scatterplot, fig.height = 5, fig.cap = "(ref:scatterplot-cap)"}
```

(ref:scatterplot-cap) Lag scatterplot in the daily calendar layout. Previous hour's count is plotted against each hour's count at Flinders Street Station to demonstrate the autocorrelation at lag 1. The correlation between them is more consistent on non-work days than work days.

@@

```{r boxplot, fig.height = 6.5, fig.cap = "(ref:boxplot-cap)"}
```

(ref:boxplot-cap) Side-by-side boxplots of hourly counts for all the 43 sensors in December 2016, with the loess smooth line superimposed on each day. It shows the hourly distribution in the city as a whole. There is one sensor attracting a larger number of people on New Year's Eve than the rest.

@@

```{r chn-embedded, fig.cap = "(ref:chn-cap)"}
include_graphics("figure/chn-1-embedded.pdf")
```

(ref:chn-cap) The same plot as Figure \ref{fig:boxplot}, but with the month and week day labels in Chinese. It demonstrates the natural support for languages other than English.

@@

```{r write-bib}
write_bib(.packages(), 'packages.bib')
```

>>>>

---
title: Calendar-based graphics for visualizing people's daily schedules
author:
  - name: Earo Wang
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics,
    - Monash University, VIC 3800
    - Australia
    email: earo.wang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics,
    - Monash University, VIC 3800
    - Australia
    email:  dicook@monash.edu
  - name: Rob J Hyndman
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics,
    - Monash University, VIC 3800
    - Australia
    email:  rob.hyndman@monash.edu
abstract: >
  Calendars are broadly used in society to display temporal information, and events. This paper describes a new R package with functionality to organize and display temporal data, collected on sub-daily resolution, into a calendar layout. The function \code{frame\_calendar} uses linear algebra on the date variable to restructure data into a format lending itself to calendar layouts. The user can apply the grammar of graphics to create plots inside each calendar cell, and thus the displays synchronize neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior in Melbourne, Australia, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.
output: 
  rticles::rjournal_article:
  includes:
    in_header: preamble.tex
bibliography: ["references.bib", "packages.bib"]
---

@@

```{r ped-map, fig.width = 4, out.width = "0.55\\linewidth", fig.cap = "Map of the Melbourne city area with purple dots indicating sensor locations."}
```

```{r time-series-plot, fig.height = 6, fig.cap = "Time series plots showing the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne. Colored by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distinctive across these 3 sensors. There is an eye-catching spike which occurred at the State Library, caused by the annual White Night event on 20th of February."}
```

```{r facet-time, fig.height = 6, fig.cap = "Hourly pedestrian counts for 2016 faceted by sensors and days of the week using lines. It features at least two types of seasons---time of day and day of week---across all the sensors. The White Night effect on increasing pedestrian counts can be seen on Saturday in February at the State Library."}
```

@@

We propose a new algorithm to go beyond the calendar-based heatmap. The approach is developed with three conditions in mind: (1) to display time-of-day variation in addition to longer temporal components such as day-of-week and day-of-year; (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout; (3) to enable overlaying plots consisting of multiple time series. The proposed algorithm has been implemented in the \code{frame\_calendar} function in the \pkg{sugrrants} package using R.

The remainder of the paper is organized as follows. Section \ref{sec:algorithm} demonstrates the construction of the calendar layout in depth. Section \ref{sec:opt} lists and describes the options that come with the \code{frame\_calendar} function. Section \ref{sec:examples} presents some variations of its usage. Section \ref{sec:discussion} discusses the advantages and disadvantages of the method.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the monthly calendar over the year 2016. This is achieved by the \code{frame\_calendar} function, which computes the coordinates on the calendar for the input data variables. These can then be plotted using the usual \pkg{ggplot2} package [@R-ggplot2] functions. All of the grammar of graphics [@wilkinson2006grammar; @wickham2009ggplot2] can be applied.

```{r flinders-2016, fig.height = 6.5, fig.cap = "The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of the data into a $3 \times 4$ monthly grid represents all the traffic in 2016. The disparities between week day and weekend along with public holiday are immediately apparent."}
```

@@

```{r month-diagram, out.width = "360pt", out.height = "250pt", fig.cap = "Illustration of the indexing layout for cells in a month, where $k$ is day of the week, $g$ is day of the month, $(i, j)$ indicates grid position."}
include_graphics("img/month.pdf")
```

@@

```{r year-diagram, out.width = "360pt", out.height = "250pt", fig.cap = "Illustration of the indexing layout for months of one year, where $M$ and $N$ indicate number of rows and columns, $b$ is a space parameter separating cells."}
include_graphics("img/year-diagram.pdf")
```

@@

The algorithm has several optional parameters that modify the layout, direction of display, scales, plot size and switching to polar coordinates. These are accessible to the user by the inputs to the function  \code{frame\_calendar}:

@@

```{r flinders-polar, fig.height = 6.5, fig.cap = "Figure \\ref{fig:flinders-2016} in circular layout, which is referred to as star plots. The daily periodicity on work days are clearly visible."}
```

@@

By default, global scaling is done for values in each plot, with the global minimum and maximum used to fit values into each cell. If the emphasis is comparing trend rather than magnitude, it is useful to scale locally. For temporal data this would harness the temporal components. The choices include: free scale within each cell (\code{free}), cells derived from the same day of the week (\code{free\_wday}), or cells from the same day of the month (\code{free\_mday}). The scaling allows for the comparisons of absolute or relative values, and the emphasis of different temporal variations.

@@

The \code{free\_wday} scales each week day together. It can be useful to comparing trends across week days, allowing relative patterns for weekends versus week days to be examined. Similarly, the \code{free\_mday} uses free scaling for any day within a given month.

```{r flinders-free, fig.height = 6.5, fig.cap = "Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive, however it is impossible to compare the size of peaks between days."}
```

@@

```{r overlay, fig.height = 7, fig.cap = "Overlaying line graphs of the 3 sensors in the monthly calendar. Flagstaff station is not as busy as the other two on non-work days."}
```

@@

```{r facet, fig.height = 11, fig.cap = "Line charts, embedded in the $6 \times 2$ monthly calendar, colored and faceted by the 3 sensors. The variations of an individual sensor are emphasised, and the shapes can be compared across the cells and sensors."}
```

@@

The \code{frame\_calendar} function is not constrained to line plots. The full range of plotting capabilities in \pkg{ggplot2} is essentially available. Figure \ref{fig:scatterplot} shows a lag scatterplot at Flinders Street Station, where the current hourly count is assigned to the \code{x} argument and the lagged hourly count to the \code{y} argument. This figure is organized in the daily calendar layout. Figure \ref{fig:scatterplot} indicates two primary patterns, strong autocorrelation on weekends, and weaker autocorrelation on work days. At the higher counts, on week days, the next hour sees possibly substantial increase or decrease in counts, essentially revealing a bimodal distribution of consecutive counts, as supported by Figure \ref{fig:flinders-2016}.

```{r scatterplot, fig.height = 5, fig.cap = "Lag scatterplot in the daily calendar layout. Previous hour's count is plotted against each hour's count at Flinders Street Station to demonstrate the autocorrelation at lag 1. The correlation between them is more consistent on non-work days than work days."}
```

@@

```{r boxplot, fig.height = 6.5, fig.cap = "Side-by-side boxplots of hourly counts for all the 43 sensors in December 2016, with the loess smooth line superimposed on each day. It shows the hourly distribution in the city as a whole. There is one sensor attracting a larger number of people on New Year's Eve than the rest."}
```

@@

```{r chn-embedded, fig.cap = "The same plot as Figure \\ref{fig:boxplot}, but with the month and week day labels in Chinese. It demonstrates the natural support for languages other than English."}
include_graphics("img/chn-1-embedded.pdf")
```

@@

```{r write-bib}
write_bib(c("ggplot2", "tidyverse", "sugrrants", "geofacet", .packages()), 'packages.bib')
```
